Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex
discussion

\begin{figure*}[tb]
  \hspace{-2.8em}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$. The horizontal axis measures the bits used to describe the model, the vertical axis measures the number of bits used to describe the noise. The smallest possible effective description, denoted by a black diagonal, is $C^\M(x)$. We accept all representations that are close to this optimum, the threshold being represented by a dashed line. The size of the smallest model within the slack is the sophistication of the data. (right) The same image, after a constant perturbation in te model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

For any string $x$, there may be many different two-part representations. The total code length can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:diagram} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, For most of these definitions, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

@@

\section{Inefficient indices}
\label{section:indices}
As mentioned in the introduction, the definition of sophistication is sometimes obtained by opening up the definition of Kolmogorov complexity, and measuring the length of the index part of that two-part code. We call such a sophistication \emph{index sophistication}. Examples are Koppel and Atlan's works \cite{koppelSoph1988,koppel1991almost},  and several of its variants \cite{antunes2009sophistication,antunes2013sophistication}.

In these works, no constraints are placed on the used numbering. This makes these approaches vulnerable to the so-called \emph{nickname problem}: the number of bits required to specify a model becomes very dependent on the numbering. This problem is severe: any index sophistication without requirements on the numbering is either bounded, or not invariant:

@@

\begin{definition}[\cite{grunwald2004shannon,vitanyi2004meaningful}]
The complexity of a computable function is defined by \label{definition:model-complexity}
\[
C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\phi_i=f\}\p
\]
\end{definition}
We show in the appendix (Lemma~\ref{lemma:invariance}) that this definition is invariant, so that the Kolmogorov complexity is an objective property of a function, as it is of a binary string. There are two ways to use this notion of model complexity for more robust attempts to define sophistication. Confusingly, both approaches are used in the literature.
\begin{enumerate}
  \item We maintain that the representation witnessing the Kolmogorov complexity determines the sophistication, but impose the additional constraint that an efficient acceptable numbering is used. We construct such an efficient index, called a \emph{faithful} index, below. This is the approach taken by Adriaans \cite{adriaans2012facticity}.
  \item On the other hand, rather than distinguishing the two components of the code \emph{within} the definition of Kolmogorov complexity, we can also use $K$ to measure the complexity of the \emph{model} (as in Definition~\ref{definition:model-complexity}), which is then the first part of a two-part code describing the data. With this approach, the total code length is $K(\phi_i)+|y|$, with $\phi_i(y)=x$. In this approach, we can construct everything on the basis of a single acceptable numbering, without additional constraints. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information}.
\end{enumerate}
We first construct an efficient numbering: we will call a numbering $\psi$ \emph{faithful} if for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$. This definition is a refinement of \cite[Definition~10]{adriaans2012facticity}.

@@

So why should we consider this intuition reasonable at all? The first assumption of sophistication is that high compressibility and pure randomness are objectively `uninteresting'. A more sophisticated signal, such as a television broadcast must contain, in some objective sense, more meaningful information. Whether this intuition seems reasonable likely differs per person \sdr{weak}, but sophistication goes one step further: it also suggests that we can objectively separate the information in a string into random and structured. \sdr{I think here we should sell the intuition better before we tear it down.}

To see whether this intuition is valid; imagine that we sample from the universal distribution by feeding random bits to a universal Turing machine, machine $u$ in the $\phi$-enumeration. If this UTM is constructed in the conventional manner, the initial bits will determine the index of a partial computable function $i$, and the rest will be the input $y$ to that machine. What is the characteristic model of the resulting data? Certainly, if we have to judge based only on the data we cannot exclude the possibility that the data was sampled from $\phi_u$: after all, it was.  On the other hand, neither can we exclude the possibility that it was generated by $\phi_i$, as again, it was!

Although sampling perspectives are discouraged by proponents of algorithmic statistics, this statistical sanity check illustrates why it seems unjustified to assume that, based on only the data, all but one model can always be disqualified. \sdr{Hier stond: There is no reason to assume that a string with high sophistication was sampled from a complex model, so how can a complex model be intrinsic to the data? \footnotemark}

\footnotetext{Note that computational depth does not have this problem. Only deep sources produce deep strings.}

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. The candidate set for this string contains various models from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal models says that it is `something computable'. A more specific model might say that it is `an image'. Even more specific would be a painting, a Monet, or specifically the painting Impression of a sunrise. These are the models we imagine we may find in the candidate set. Can we say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? We see no intuitive reason why such a distinction should be possible without making any further assumptions.

Of course, \sdr{Actually this is not trivial at all} Kolmogorov complexity allows us to say that it is unlikely to be a Jackson Pollock, or a piece of music. And, if we were to get a second sample from the same source, we \emph{could} answer the question: we would receive either the same painting again, or perhaps another Monet. But, a single sample does not give us enough evidence to choose a single model. \sdr{Volgende zin weer zwak, de metafoor wordt backed up door ons voorbeeld hierboven dus is behoorlijk hard} Ultimately, this is just a metaphor, but since the notion of sophistication is driven by intuition, we feel that an intuition to the contrary should serve to sharpen the debate.

For this reason, we take a skeptical view of sophistication. We can offer the following insight, to back up our skepticism. As we have seen, candidates within a constant of the optimal representation can occur all over the spectrum, from small to large models. A small change in construction, such as a different acceptable numbering, or a change in threshold function, can arbitrarily push models into or out of the candidate set. Whereas in Kolmogorov complexity, constant changes have constant effects on the outcome, here constant changes can cause large changes in the outcome. \sdr{Not sufficiently clear} While no authors have touched on the issue of invariance, it is related to the issue of instability \cite{antunes2013sophistication,vereshchagin2013algorithmic}. We consider invariance the first criterion by which any proposal must be judged, with the burden of proof resting with the proposer, rather than the critic.

@@

\sdr{Something about what is useful about the theory}
\sdr{A picture of the structure function (perhaps sooner as well)}
\sdr{This has become a discussion section. It does not recap. That's a bit unusual.}

\section{Conclusion}
As we have seen, the idea behind sophistication has a rich history. It has been proposed many times, by many different authors. This tells us two things. First, there is a strong intuition behind sophistication, shared by many respected authors, and second, it is not easy to get the definition exactly right.

@@

\ineffprefix*
\begin{proof}
Let $\bar\imath y$ be any two-part representation for the data $x$, i.e. $\phi_i(y)=x$. Then construct an alternative two-part representation $\bar vi^* y$, where $i^*$ is the shortest $\psi_u$-program for $i$ such that $\phi_v(i^* y)=\phi_{\psi_u(i^*)}(y) = \phi_i(y)=x$. We compare the lengths of these two representations. Note that $|i^*|=K^\psi(i)$. Therefore,
\[
|\bar\imath p|-|\bar v i^* y| = |\bar\imath|-|\bar v| - |i^*| = |\bar\imath|-|\bar v|-K(i).
\]
By assumption there must be an $i_0$ such that the above expression is positive for all $i>i_0$. From this $i_0$ onwards, the second representation (using $v$) will have a shorter code length, so $\bar\imath p$ cannot achieve the minimum in the definition of the sophistication. Consequently, the sophistication is bounded by $\s(x)<|\overline{\imath_0}|$ for all $x$. 
\end{proof}

>>>>

\begin{figure*}[tb]
  \hspace{-2.8em}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication.pdf}
  \end{minipage}
  \begin{minipage}{0.55\textwidth}
     \includegraphics[width=\textwidth]{./img/sophistication-jump.pdf}
  \end{minipage}
  \caption{\small (left) Two-part representations of $x$. The horizontal axis measures the bits used to describe the model, the vertical axis measures the number of bits used to describe the noise. The Kolmogorov complexity $C^\M(x)$, appearing as a black diagonal, provides a lower bound. We accept all representations that are close to this optimum, the threshold being represented by a dashed line. The size of the smallest model within the slack is the sophistication of the data. (right) The same image, after a constant perturbation in te model complexity caused by a change in numbering.}
  \label{fig:diagram}
\end{figure*}

For any string $x$, there may be many different two-part representations. The total codelength can never be less than the Kolmogorov complexity, but it can come close. The balance between the two parts may vary; Figure~\ref{fig:diagram} provides an illustration. The various proposals then define sophistication in terms of the length of the structure part of this two-part code for one of the representations that (almost) achieves the Kolmogorov complexity. However, For most of these definitions, we can prove that they fail one or more of the conditions above. Vit\'anyi's definition \cite{vitanyi2004meaningful} is the one exception that we cannot \emph{prove} to conflict with our requirements, but we can show that only strings that require an enormous amount of processing to construct, yield a large sophistication.

@@

\section{Inefficient indices}
\label{section:indices}
In some works, the definition of sophistication is obtained by opening up the definition of Kolmogorov complexity, and measuring the length of the index part of that two-part code. We call such a sophistication \emph{index sophistication}. Examples are Koppel and Atlan's works \cite{koppelSoph1988,koppel1991almost},  and several of its variants \cite{antunes2009sophistication,antunes2013sophistication}.

In these works, no constraints are placed on the used numbering. This makes these approaches vulnerable to the so-called \emph{nickname problem}: the number of bits required to specify a model becomes very dependent on the numbering. This problem is severe: \emph{any} index sophistication without requirements on the numbering is either bounded, or not invariant:

@@

\begin{definition}[\cite{grunwald2004shannon,vitanyi2004meaningful}]
The complexity of a computable function is defined by \label{definition:model-complexity}
\[
C^{\M,\psi}(f) = \min\{C^{\M,\psi}(i):\phi_i=f\}\p
\]
\end{definition}
We show in the appendix (Lemma~\ref{lemma:invariance}) that this definition is invariant, so that the Kolmogorov complexity is an objective property of a function, as it is of a binary string. There are two ways to use this notion of model complexity for more robust attempts to define sophistication. Confusingly, both approaches are used in the literature.
\begin{enumerate}
  \item We maintain that the representation witnessing the Kolmogorov complexity determines the sophistication, but impose the additional constraint that an efficient acceptable numbering is used. We construct such an efficient index, called a \emph{faithful} index, below. This is the approach taken by Adriaans \cite{adriaans2012facticity}.
  \item On the other hand, rather than distinguishing the two components of the code \emph{within} the definition of Kolmogorov complexity, we can also use $K$ to measure the complexity of the \emph{model} (as in Definition~\ref{definition:model-complexity}), which is then the first part of a two-part code describing the data. With this approach, the total codelength is $K(\phi_i)+|y|$, with $\phi_i(y)=x$. In this approach, we can construct everything on the basis of a single acceptable numbering, without additional constraints. This approach is used in  \cite{cover1985kolmogorov,gacs2001algorithmic,vitanyi2004meaningful,gellmann1996information}.
\end{enumerate}
We first construct an efficient numbering: we will call a numbering $\psi$ \emph{faithful} if for all $i$ there is a $j$ such that $\psi_i=\psi_j$ and $|j|\le C^{\C}(\psi_j)+c$, for some constant $c$. This definition is a refinement of \cite[Definition~10]{adriaans2012facticity}.

@@

We have criticized existing measures of the sophistication of a binary string. Our results show that there are technical problems with each of them. But that does not in itself mean that it should be impossible to come up with a sound measure. The intuition throughout the literature, starting with Kolmogorov's proposal of the structure function, appears to have been that it should be possible to identify noise, and distinguish it from structural information, by its property of being incompressible.

This idea is bold, but not necessarily unreasonable. Nevertheless, in this section we offer the opinion that such a clean-cut separation \emph{cannot} be made to work. We provide two arguments to support this belief.

For the first argument, we take a generative perspective. If data are generated by a computable probability distribution, then a sophistication could be called \emph{consistent} if, given a large enough sample, it accurately reflects the complexity of that distribution. While in general we do not wish to make any assumptions about the origin of the data, it can make sense to do so as a sanity check, as we will do here. Suppose that the data are generated by a universal distribution, e.g. by feeding random bits to a universal partial computable function. Let $u$ be the index of the function in the $\phi$-enumeration with the property that $\phi_u(\bar\imath y)=\phi_i(y)$. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of data that is generated in this fashion?

Certainly, if we have to judge based only on the data, we cannot exclude the possibility that the data was sampled from $\phi_u$: after all, it was.  On the other hand, neither can we exclude the possibility that it was generated by $\phi_i$, as again, it was! For this reason it seems unjustified to assume that, based on only the data, all but one model can always be disqualified.

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. There are many good models for this string, ranging from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal models says that it is `something computable'. A more specific model might say that it is `an image'. Even more specific would be a painting, a Monet, or specifically the painting Impression of a Sunrise. A sound measure of Sophistication should be able to select one of these models as the proper representation of structure in the data, and disqualify the others as either over- or underfitting. But how should we be able to say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? Intuitively such distinctions require further assumptions.

The second argument why we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Figure~\ref{fig:diagram} provides an illustration. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in the figure. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. \emph{However}, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps do not occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that \emph{part} of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=\min\{|y|:\phi_i(y)=x\}-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data --- if $\phi_i$ truly generated the data, then the probability that the random deficiency is more than $k$ bits is less than $2^{-k}$ (proof omitted). But fundamental problems arise as soon as hard cut-offs are introduced. We also wish to point out that a lot of measures taken in the literature to address problems with sophistication, such as restricting the model class or introducing model penalties, are no longer necessary as soon as such hard cut-offs are avoided. What remains is a simple and elegant theory that can disprove most models, leaving not one, but a limited number of interesting candidates.

\section{Conclusion}
As we have seen, the idea behind sophistication has a rich history. It has been proposed many times, by many different authors. This tells us two things. First, there is a strong intuition behind sophistication, shared by many respected authors, and second, it is not easy to get the definition exactly right.
 While no authors have touched on the issue of invariance, it is related to the issue of instability \cite{antunes2013sophistication,vereshchagin2013algorithmic}. We consider invariance a first criterion that any proposal must satisfy, , with the burden of proof resting with the proposer, rather than the critic.

@@

@@

\ineffprefix*
\begin{proof}
Let $\bar\imath y$ be any two-part representation for the data $x$, i.e. $\phi_i(y)=x$. Then construct an alternative two-part representation $\bar vi^* y$, where $i^*$ is the shortest $\psi_u$-program for $i$ such that $\phi_v(i^* y)=\phi_{\psi_u(i^*)}(y) = \phi_i(y)=x$. We compare the lengths of these two representations. Note that $|i^*|=K^\psi(i)$. Therefore,
\[
|\bar\imath p|-|\bar v i^* y| = |\bar\imath|-|\bar v| - |i^*| = |\bar\imath|-|\bar v|-K(i).
\]
By assumption there must be an $i_0$ such that the above expression is positive for all $i>i_0$. From this $i_0$ onwards, the second representation (using $v$) will have a shorter codelength, so $\bar\imath p$ cannot achieve the minimum in the definition of the sophistication. Consequently, the sophistication is bounded by $\s(x)<|\overline{\imath_0}|$ for all $x$. 
\end{proof}
